Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2

### 2023-08-23 go3.Rmd
makefile and Michelle's comments
Makefile now, all code external, all data loaded from gloor github data repository via URL begun supplemental

One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data from different samples. A large number of normalizations were developed that depended on the data source. These include rarefaction[@Hughes:2005tu], proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a], and the centered log ratio (CLR)[@aitchison1982]. With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.

Recently, [,@nixon2023scale] showed that the denominator used to normalize \(Y^{\parallel}_{n}\) was an estimate of the scale of the system. In the context of compositional data analysis, this group proved that the geometric mean of \(Y^{\parallel}_{n}\) was a biased estimate of \(W^{\perp}_n\). This realization opened the possibility that \(Y^{\perp}_{n}\) could be made a less biased estimate by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated. In terms of information theory, the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(I\) (supplemental figure XX), suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional. This makes intuitive sense because  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim I_n\).

The realization that a strong assumption about scale is built into each denominator explains why HTS data are notoriously fickle as analysis of the same dataset with different workflows, tools and assumptions giving widely different results when benchmarked[@Thorsen:2016aa;@Quinn:2018aa;@Gardner:2019aa]. This can be understood as the result of unacknowledged bias about the scale estimate which inevitably leads to large Type 1 error rates as sample sizes increase [@Gustafson2015]. Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be robust to differential abundance analysis even when taking scale into account.

@@

```{r dispersion, echo=F,warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2. The expected clr value for each transcipt are plotted vs. the expected dispersion. In panels B and C the amount of dispersion reaches a minimum at moderate  values. The red line in each panel shows the loess line of fit to the mid-point of the distributions. The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The dashed grey line in panels B and C represent the maximum of the lowest 10% dispersion." }
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

yst <- read.table('~/Documents/0_git/datasets/transcriptome.txt', header=T, 
  row.names=1)
# remove the one gene with 0 reads

@@

# DESeq2
dds <- DESeqDataSetFromMatrix(countData = yst,
          colData = coldata, design= ~ conds)
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="conds_W_vs_S")
set.seed(2023)
#ALDEx2
x <- aldex.clr(yst, conds, gamma=1e-3,verbose=F)
x.e <- aldex.effect(x, include.sample.summary=T, verbose=F)
x.t <- aldex.ttest(x, verbose=F)
set.seed(2023)
x.s <- aldex.clr(yst, conds, gamma=1, verbose=F)
x.s.e <- aldex.effect(x.s, include.sample.summary=T, verbose=F)
x.s.t <- aldex.ttest(x.s, verbose=F)

sig.des <- which(res@listData$padj < 0.01)
sig.ald <- which(x.t$we.eBH < 0.01)
sig.s.ald <- which(x.s.t$we.eBH < 0.01)

@@

l.df <- data.frame(x.e$rab.all, x.e$diff.win)
plot(x.e$rab.all, x.e$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(dispersion)')
#points(x.e$rab.all[sig.ald], x.e$diff.win[sig.ald], col='orange', cex=0.4)
lines(lowess(l.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
```

Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data[@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution. This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes. This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation[@Taniguchi:2010aa]. Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible. To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives. The work of [,@nixon2023scale] suggests that this is because the biased estimate of \(Y^{\perp}\) that is  calculated leads to a false estimate of the scale. Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a FDR of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts. Clearly, the necessary assumption of most features being invariant is not justified. Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.t$we.eBH < 0.01 & abs(x.e$diff.btw) > 1.4))` for ALDEx2.

@@

plot(x.e$diff.win, x.e$diff.btw, col=rgb(0,0,0,0.1), 
   xlab='Dispersion', ylab='log2 Difference')
title('B: ALDEx2 effect', adj=0, line= 0.8)
points(x.e$diff.win[sig.ald], x.e$diff.btw[sig.ald], 
  col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')
points(x.e$diff.win[sig.s.ald], x.e$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.6), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.e$diff.win, x.s.e$diff.btw, col=rgb(0,0,0,0.1),
   xlab='Dispersion', ylab='log2 Difference', xlim=c(0.1,5))
title('C: ALDEx2 scaled effect', adj=0, line= 0.8)
points(x.s.e$diff.win[sig.s.ald], x.s.e$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

@@

plot(x.e$diff.btw, -1*log10(x.t$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('E: ALDEx2 volcano', adj=0, line= 0.8)
points(x.e$diff.btw[sig.ald], -1*log10(x.t$we.eBH[sig.ald]  +1e-70), 
  col=rgb(1,.66,0,.3), cex=0.5, pch=19)
points(x.e$diff.btw[sig.s.ald], -1*log10(x.t$we.eBH[sig.s.ald]  +1e-70), 
  col=rgb(1,0,0,1), cex=0.5, pch=19)
abline(v=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.e$diff.btw, -1*log10(x.s.t$we.eBH +1e-15), col=rgb(0,0,0,0.1),
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('F: ALDEx2 scaled volcano', adj=0, line= 0.8)

points(x.s.e$diff.btw[sig.s.ald], -1*log10(x.s.t$we.eBH[sig.s.ald]+1e-15),
  col=rgb(1,0,0,1), cex=0.5, pch=19)
    abline(v=c(-1.4,1.4), lty=2, col='grey')

@@

As shown in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2. We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups. In the most extreme cases transcripts with near 0 difference have a low FDR. This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines. A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts. This results begs the question, "why bother with the signficance test?".

@@

```{r makeScaleMatrix, echo=F}
aldex.makeScaleMatrix <- function(gamma, diff, conditions){
  ## new scale model
  # mu1 is the base relative variance
  # mu2 is the other relative variance
  # gamma is the dispersion paramenter
  gamma=gamma
  mu1 = 1 # set base to 1
  mu2 = 1 + diff # set other to adjust this until the housekeeping is centred

  # note: it is the log2 difference between mu1 and mu2 that is key here
  # eg; mu1=1, mu2=1.15 is equivalent to mu1=4, mu2=4.6
  # log2(1)=0, log2(1.15)~0.2; log2(4)=2, log2(4.6)~2.2

  mu.vec <- gsub(levels(factor(conditions))[1], log2(mu1), conditions)
  mu.vec <- as.numeric(gsub(levels(factor(conditions))[2], log2(mu2), mu.vec))
  return(t( sapply(mu.vec, FUN = function(mu) rlnorm(128, mu, gamma)) ))
}
```

@@

set.seed(2023)
load('~/Documents/0_git/projects/metatranscriptome/Rdata/ko.both.all.Rda')

# make a vector of conditions
conds.K0 <- c(rep('H',8), rep('B',14), rep('B',14), rep('H', 8))

xt <- aldex.clr(ko.both.all, conds.K0)
xt.e <- aldex.effect(xt, include.sample.summary=T)
xt.t <- aldex.ttest(xt)
xt.all <- cbind(xt.e, xt.t)

## single gamma model
xg <- aldex.clr(ko.both.all, conds.K0, gamma=1)
xg.e <- aldex.effect(xg)
xg.t <- aldex.ttest(xg)
xg.all <- cbind(xg.e, xg.t)

mu.mod <- aldex.makeScaleMatrix(0.75, .15, conds.K0)

xt.m <- aldex.clr(ko.both.all, conds.K0, gamma=mu.mod)
xt.m.e <- aldex.effect(xt.m, include.sample.summary=T)
xt.m.t <- aldex.ttest(xt.m)
xt.m.all <- cbind(xt.m.e, xt.m.t)

hk <- rownames(xt.e)[xt.e$diff.win < 2.5 & xt.e$diff.btw > 0 & xt.e$diff.btw < 3]
hk.off <- xt.e$diff.win < 2.5 & xt.e$diff.btw > 0 & xt.e$diff.btw < 3

@@

Applying `gamma=1` as before increases the dispersion as expected, but does little to move the large number of housekeeping functions toward the midline of no difference. Nevertheless, about 50% of the housekeeping functions are no longer statistically significantly different.

Up to this point, scale uncertainty has been applied uniformly to both conditions, but the scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix. This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group. Applying a differential scale of 0.15, or 15% of the base scale now moves the housekeeping functions to the midline of no difference (Figure 3C). Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because the change in scael moves the mass of the data towards the group with the highest entropy (supplemental). Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied. These former false negatives are noted in orange in each panel. Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant. Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset.

@@

While the underlying scale of the system cannot be measured easily, the effect on analysis can by including scale uncertainty in the analysis [@nixon2023scale]. Nixon et al, showed that this can be done by including uncertainty in the denominator used for the normalization. The ALDEx2 R package is ideally suited for this since this tool builds a Bayesian posterior of the compositional component of the dataset at the outset and then conducts the analysis on that posterior. Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty. For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a log-Normal distribution.

>>>>

One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data from different samples. A large number of normalizations were developed. Rarefaction[@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a]. The centered log ratio (CLR)[@aitchison1982] approach was proposed as a unifying normalization[@fernandes:2014]. With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.

Recently, [,@nixon2023scale] showed that geometric mean of \(Y^{\parallel}_{n}\) was related to the geometric mean of \(W_n\) and so was related to \(W^{\perp}_n\). This realization opened the possibility that adding uncertainty to the geometric mean could provide a buffer of uncertainty on it that could be used to control for  the unacknowledged bias that the geometric mean was a perfect point estimate of the underlying system. The bias around  could be acknowledged by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated. In terms of information theory, the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(I\) (supplemental figure XX), suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional. This makes intuitive sense because  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim I_n\).

The realization that a strong assumption about scale is built into each denominator explains why HTS data are notoriously fickle as analysis of the same dataset with different workflows, tools and assumptions giving widely different results when benchmarked[@Thorsen:2016aa;@Quinn:2018aa;@Gardner:2019aa]. This can be understood as the result of unacknowledged bias about the scale estimate which inevitably leads to large Type 1 error rates as sample sizes increase [@Gustafson2015]. Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be called as differentially abundant even when scale was taken into account, and those features that are differentially abundant only under strictly assumed normalizations.

@@

```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2. The expected clr value for each transcipt are plotted vs. the expected dispersion. In panels B and C the amount of dispersion reaches a minimum at moderate  values. The red line in each panel shows the loess line of fit to the mid-point of the distributions. The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The dashed grey line in panels B and C represent the maximum of the lowest 10\\% dispersion." }

library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
yst <- read.table(url, header=T, row.names=1)
# remove the one gene with 0 reads

@@

# DESeq2
load('analysis/res.Rda')

#ALDEx2
load('analysis/x.all.Rda')
load('analysis/x.s.all.Rda')

# with gamma = 1
load('analysis/x.s.all.Rda')
sig.des <- which(res@listData$padj < 0.01)
sig.ald <- which(x.all$we.eBH < 0.01)
sig.s.ald <- which(x.s.all$we.eBH < 0.01)

@@

l.df <- data.frame(x.all$rab.all, x.all$diff.win)
plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(dispersion)')
#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
lines(lowess(l.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
```

Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data[@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution. This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes. This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation[@Taniguchi:2010aa]. Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible. To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion. The work of [,@nixon2023scale] suggests that this is because the biased estimate of \(Y^{\perp}\) that is  calculated leads to a false estimate of the scale. Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts. Clearly, the necessary assumption of most features being invariant is not justified. Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.all$we.eBH < 0.01 & abs(x.all$diff.btw) > 1.4))` for ALDEx2.

@@

plot(x.all$diff.win, x.all$diff.btw, col=rgb(0,0,0,0.1), 
   xlab='Dispersion', ylab='log2 Difference')
title('B: ALDEx2 effect', adj=0, line= 0.8)
points(x.all$diff.win[sig.ald], x.all$diff.btw[sig.ald], 
  col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')
points(x.all$diff.win[sig.s.ald], x.all$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.6), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.all$diff.win, x.s.all$diff.btw, col=rgb(0,0,0,0.1),
   xlab='Dispersion', ylab='log2 Difference', xlim=c(0.1,5))
title('C: ALDEx2 scaled effect', adj=0, line= 0.8)
points(x.s.all$diff.win[sig.s.ald], x.s.all$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.5), cex=0.5, pch=19)
    abline(h=c(-1.4,1.4), lty=2, col='grey')

@@

plot(x.all$diff.btw, -1*log10(x.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('E: ALDEx2 volcano', adj=0, line= 0.8)
points(x.all$diff.btw[sig.ald], -1*log10(x.all$we.eBH[sig.ald]  +1e-70), 
  col=rgb(1,.66,0,.3), cex=0.5, pch=19)
points(x.all$diff.btw[sig.s.ald], -1*log10(x.all$we.eBH[sig.s.ald]  +1e-70), 
  col=rgb(1,0,0,1), cex=0.5, pch=19)
abline(v=c(-1.4,1.4), lty=2, col='grey')

plot(x.s.all$diff.btw, -1*log10(x.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
   xlab='log2 Difference', ylab='-1 log10(p.adjust)')
title('F: ALDEx2 scaled volcano', adj=0, line= 0.8)

points(x.s.all$diff.btw[sig.s.ald], -1*log10(x.s.all$we.eBH[sig.s.ald]+1e-15),
  col=rgb(1,0,0,1), cex=0.5, pch=19)
    abline(v=c(-1.4,1.4), lty=2, col='grey')

@@

As shown in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2. We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups. In the most extreme cases transcripts with near 0 difference have a low FDR. This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines. A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts. This results begs the question, "why bother with the signficance test?". [XXXX CHECK THE DIRECTION OF SCALE IN THIS DATASET]

@@

@@

# all code for ALDEx2 and DESeq is being moved to the code/directory
# all variables recreated using the makefile

load('analysis/xt.all.Rda')
load('analysis/xg.all.Rda')
load('analysis/xt.m.all.Rda')

hk <- rownames(xt.all)[xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3]
hk.off <- xt.all$diff.win < 2.5 & xt.all$diff.btw > 0 & xt.all$diff.btw < 3

@@

Applying the default scale model of `gamma=1` as before increases the dispersion as expected, but does little to move the large number of housekeeping functions toward the midline of no difference. Nevertheless, about 50% of the housekeeping functions are no longer statistically significantly different.

Up to this point, scale uncertainty has been applied uniformly to both conditions, but a user-defined scale adjustment can be applied to each condition, or even each sample independently through a custom scale matrix. This can be done quite simply with the `aldex.makeScaleMatrix()` function which produces a matrix of scale uncertainties that are distinct for each group. Applying a per-group differential scale of 0.15, or 15% along with the base scale now moves the housekeeping functions to the midline of no difference (Figure 3C). Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant because the change in scale moves the mass of the data towards the group with the highest entropy (supplemental). Note that this identifies a significant number of functions that are differentially up in BV that were formerly classed as not different without scale, or when only a uniform  scale was applied. These former false negatives are noted in orange in each panel. Inspection of the functions shows that these are largely missing from the Lactobacillus species and so should actually be captured as differentially abundant. Thus, applying differential scale allows us to distinguish between both false positives (housekeeping functions in cyan) and false negatives (orange functions) even in a very difficult to analyze dataset. Thus, while the default scale model alone is useful and should be normally applied, using prior information about the underlying biology to develop a differential scale model can be very useful for many datasets.

@@

Since the underlying scale of the system cannot be measured easily --  methods such as flow cytometry [@Vandeputte:2017aa], spike-in probes and fluorescent in-situ hybridization [@Loven:2012aa;@yeast-absolute] have been used -- the effect on analysis can be modelled by including scale uncertainty in the analysis [@nixon2023scale]. Nixon et al, showed that this can be done by including uncertainty in the denominator used for the normalization. The ALDEx2 R package is ideally suited for this since this tool builds a Bayesian posterior of the compositional component of the dataset at the outset and then conducts the analysis on that posterior. Adding scale uncertainty can be done at the same time thus producing a posterior model that incorporates both compositional and scale uncertainty. For this, the compositional uncertainty is sampled from a Dirichlet distribution, and the scale uncertainty is sampled from a log-Normal distribution.
